---
title: "homework"
author: "Jakub Guzek"
date: "2023-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dependancies and setup

```{r load}
library(devtools)
library(ggplot2)
library(ggpubr)
library(gplots)
library(sva)
library(bladderbatch)
library(broom)
library(tidyverse)
library(data.table)
library(RColorBrewer)
library(huxtable)
```

Set up colors and themes.

```{r}
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))
my_colors = colorRampPalette(c("violet", "white", "darkorange"))(n=599)
my_colors_2 = colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(n=29)
```

## Load the bladder cancer data

Load the expression data from 57 bladder samples. Samples were processed in 5 batches. [Original study](https://cancerres.aacrjournals.org/content/canres/64/11/4040.full.pdf)

```{r}
data(bladderdata)
phenotypes <- pData(bladderEset)
expression_data <- exprs(bladderEset)
phenotypes
class(expression_data)
```

Look for N/A nad NaN values in the datasets

```{r}
print("Expression data:")
sum(is.na(expression_data))
print("Metadata:")
sum(is.na(phenotypes))
```

Plot mean vs variance

```{r}
row.variances <- apply(expression_data, 1, function(x) var(x))
row.means <- apply(expression_data, 1, function(x) mean(x))
ggplot(data.frame(row.means, row.variances), aes(x=row.means, y = row.variances)) + geom_point() + geom_smooth(method="lm", col = "lightcoral") + labs(title = "Mean vs. Variance relationship") 
```

Normalize the data

```{r}
expression_data <- expression_data[row.variances < 5, ]
expression_data <- log2(expression_data)
row.variances <- apply(expression_data, 1, function(x) var(x))
row.means <- apply(expression_data, 1, function(x) mean(x))
ggplot(data.frame(row.means, row.variances), aes(x=row.means, y = row.variances)) + geom_point() + geom_smooth(method="lm", col = "lightcoral") + labs(title = "Mean vs. Variance relationship") 
```

## Homework problem 1

Create the table

```{r}
#batch.table <- unique(phenotypes$batch) %>% map_dfc(setNames, object = phenotypes$outcome)
batch.table <- table(phenotypes$cancer, phenotypes$batch, dnn = c("Batch", "Outcome"))
batch.table <- as_huxtable(batch.table)
caption(batch.table) <- "Number of certain outcome by batch number"
bottom_border(batch.table)[1, 1:ncol(batch.table)] <- 0.5
batch.table <- set_background_color(batch.table, evens, everywhere, "gray95")
batch.table <- set_bold(batch.table, row = 1, col = everywhere)
batch.table <- set_bold(batch.table, col = 1, row = everywhere)
#quick_pdf(batch.table, file = "Guzek_problem1.pdf") 
print_screen(batch.table, file = "Guzek_problem1.pdf") 
```


## Homework problem 2

Sort the data by batches
```{r}
# Clumsy way to sort somples by batches from which they come from
ettiquetted_edf <- arrange(data.frame(phenotypes, t(expression_data)), batch)
expression_data <- t(as.matrix(arrange(data.frame(t(expression_data)), phenotypes$batch)))
expression_data[1,]
```

Prepare the cleaned dataset using ComBat

```{r}
batch <- phenotypes$batch
combat_expression_data <- ComBat(dat = expression_data,
                                 batch = batch, 
                                 mod = model.matrix(~1, data = phenotypes),
                                 par.prior = TRUE,
                                 prior.plots = TRUE)

```
Take a look at data cleaned by ComBat

```{r}
head(combat_expression_data[,1:6])
```

Create the heatmaps


```{r}
#png("Guzek_problem2a.png", width = 900, height = 900)
heatmap.2(as.matrix(expression_data),
          main = "Bladder Cancer Data Sorted by Bath, Before ComBat",
          notecol = "black",
          Colv = FALSE,
          density.info = "none",
          trace = "none",
          margins = c(12,9),
          col = my_colors,
          dendrogram = "none",
          scale = "row",
          ColSideColors = palette("Set2")[as.factor(ettiquetted_edf$batch)]
)
#dev.off()
```

```{r}
#png("Guzek_problem2b.png", width = 900, height = 900)
heatmap.2(as.matrix(combat_expression_data),
          main = "Bladder Cancer Data Sorted by Batch, After Combat",
          notecol = "black",
          Colv = FALSE,
          density.info = "none",
          trace = "none",
          margins = c(12,9),
          col = my_colors,
          dendrogram = "none",
          scale = "row",
          ColSideColors = palette("Set2")[as.factor(ettiquetted_edf$batch)]
)
#dev.off()
```


# Homework problem 3

```{r}
correlation.matrix <- apply(expression_data, 2, cor, expression_data)
rownames(correlation.matrix) <- colnames(correlation.matrix)
#png("Guzek_problem3a.png", width = 1000, height = 1000)
heatmap.2(correlation.matrix,
          main = "Pairwise Pearson Comparison of Expression Data, Samples Grouped by Batch",
          notecol = "black",
          density.info = "none",
          Colv = F,
          Rowv = F,
          trace = "none",
          colsep = 1:ncol(correlation.matrix),
          rowsep = 1:ncol(correlation.matrix),
          sepcolor = "gray",
          margins = c(9, 6),
          dendrogram = "none",
          col = my_colors_2,
          RowSideColors = palette("Set2")[as.factor(ettiquetted_edf$batch)],
          ColSideColors =  palette("Set2")[as.factor(ettiquetted_edf$batch)]
)
#dev.off()

ettiquetted_edf = arrange(ettiquetted_edf, outcome)
expression_data = t(as.matrix(arrange(data.frame(t(expression_data)), phenotypes$outcome)))

correlation.matrix <- apply(expression_data, 2, cor, expression_data)
rownames(correlation.matrix) <- colnames(correlation.matrix)
#png("Guzek_problem3b.png", width = 1000, height = 1000)
heatmap.2(correlation.matrix,
          main = "Pairwise Pearson Comparison of Expression Data, Samples Grouped by Outcome",
          notecol = "black",
          density.info = "none",
          Colv = F,
          Rowv = F,
          trace = "none",
          colsep = 1:ncol(correlation.matrix),
          rowsep = 1:ncol(correlation.matrix),
          sepcolor = "gray",
          margins = c(9, 6),
          dendrogram = "none",
          col = my_colors_2,
          RowSideColors = palette("Set2")[as.factor(ettiquetted_edf$outcome)],
          ColSideColors =  palette("Set2")[as.factor(ettiquetted_edf$outcome)]
)
#dev.off()
```

## Homework problem 4

Import the bottomly datasest

```{r}
if (file_test("-f", "./bottomly.Rdata")) {
  load(file = "./bottomly.Rdata")
} else {
  con <- url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bottomly_eset.RData")
  load(file = con)
  close(con)
  save(bottomly.eset, file = "bottomly.Rdata")
}
```

Get the expression data

```{r}
bottomly_edata <- as.matrix(exprs(bottomly.eset))
bottomly_edata <- bottomly_edata[rowMeans(bottomly_edata) > 10, ]
bottomly_edata <- log2(as.matrix(bottomly_edata) + 1)
bottomly_metadata <- pData(bottomly.eset)
dim(bottomly_edata)
bottomly_edata[1:10,1:6]
bottomly_metadata
```

```{r}
mod = lm(t(bottomly_edata) ~ as.factor(bottomly_metadata$strain) + as.factor(bottomly_metadata$experiment.number))
tidy_mod <- tidy(mod)
tidy_mod
names(tidy_mod)
```

```{r}
png("Guzek_problem4a.png", width = 800, height = 700)
coeff = ggplot(tidy_mod %>% filter(term == "as.factor(bottomly_metadata$strain)DBA/2J")) + geom_histogram(aes(x=estimate), bins = 100, fill = "darkorange")
p_val = ggplot(tidy_mod %>% filter(term == "as.factor(bottomly_metadata$strain)DBA/2J")) + geom_histogram(aes(x=p.value), bins = 100, fill="darkorange")
figure <- ggarrange(coeff, p_val, labels = c("Histogram of coefficients of uncorrected data", "Histogram of p-values of uncorrected data"), ncol = 1, nrow = 2) + labs(title = "Guzek_problem4a")
annotate_figure(figure, top = text_grob("Guzek_problem4a"))
dev.off()
figure
```

```{r}
bottomly_batch <- bottomly_metadata$experiment.number
combat_bottomly_edata <- ComBat(dat = bottomly_edata,
                                batch = bottomly_metadata$experiment.number,
                                mod = model.matrix(~1, data = bottomly_metadata),
                                par.prior = TRUE,
                                prior.plots = TRUE
)
```

Fit linear model on combat corrected bottomly data.

```{r}
mod_combat <- lm(t(combat_bottomly_edata) ~ as.factor(bottomly_metadata$strain))
tidy_mod_combat <- tidy(mod_combat)
tidy_mod_combat
names(tidy_mod_combat)
```

```{r}
png("Guzek_problem4b.png", width = 800, height = 700)
coeff = ggplot(tidy_mod_combat %>% filter(term == "as.factor(bottomly_metadata$strain)DBA/2J")) + geom_histogram(aes(x=estimate), bins = 100, fill = "darkorange")
p_val = ggplot(tidy_mod_combat %>% filter(term == "as.factor(bottomly_metadata$strain)DBA/2J")) + geom_histogram(aes(x=p.value), bins = 100, fill="darkorange")
figure <- ggarrange(coeff, p_val, labels = c("Histogram of coeffecients of corrected data", "Histogram of p-values of corrected data"), ncol = 1, nrow = 2) + labs(title = "Guzek_problem4b")
annotate_figure(figure, top = text_grob("Guzek_problem4b"))
dev.off()
figure
```

Compare the coefficients of uncorrected and corrected data

```{r}
est_compare <- tibble(
  Uncorrected = tidy_mod %>% filter(term == "as.factor(bottomly_metadata$strain)DBA/2J") %>% select("estimate") %>% unlist,
  Corrected = tidy_mod_combat %>% filter(term == "as.factor(bottomly_metadata$strain)DBA/2J") %>% select("estimate") %>% unlist)

png("Guzek_problem4c.png", width = 800, height = 800)
figure <- ggplot(est_compare, aes(x = Uncorrected, y = Corrected)) +
     geom_point(col = "darkgrey", alpha = .5, size = .5) + geom_abline(intercept = 0, slope = 1, col = "darkorange") + geom_smooth(method = "lm") + labs(title = "Guzek_problem4c")
figure
dev.off()
figure
```

## Homework problem 5

Find a dimension of surrogate variables

```{r}
mod = model.matrix(~as.factor(strain), data = bottomly_metadata)
mod0 = model.matrix(~1, data = bottomly_metadata)
num.sv(bottomly_edata, mod, method = "be")
num.sv(bottomly_edata, mod, method = "leek")
sva.output <- sva(bottomly_edata, mod, mod0, n.sv=num.sv(bottomly_edata, mod, method = "leek"))
```

Take a look at the data

```{r}
head(sva.output$sv)
```

Look at the relation between the surrogate variables and experiment number 

```{r}
summary(lm(sva.output$sv ~ bottomly_metadata$experiment.number))
```

Visualize the SVs

```{r}
sva.batch <- tibble(SV1 = sva.output$sv[,1],
                  SV2 = sva.output$sv[,2],
                  batch = as.factor(bottomly_metadata$experiment.number),
                  strain = as.factor(bottomly_metadata$strain),
                  lane = as.factor(bottomly_metadata$lane.number))
```

```{r}
batch.plot <- ggplot(sva.batch) + geom_point(aes(x = SV1, y = SV2, col = batch))
batch.plot
```

```{r}
strain.plot <- ggplot(sva.batch) + geom_point(aes(x = SV1, y = SV2, col = strain))
strain.plot
```

```{r}
lane.plot <- ggplot(sva.batch) + geom_point(aes(x = SV1, y = SV2, col = lane))
lane.plot
```


```{r}
sva.batch <- tibble(SV1 = sva.output$sv[,1],
                    SV2 = sva.output$sv[,2],
                    batch = as.factor(bottomly_metadata$experiment.number))
sva.batch.gather <- gather(sva.batch, "sv", "value", -batch)

ggplot(sva.batch.gather) + geom_violin(aes(x = batch, y = value)) + facet_wrap(~sv, ncol = 1) + geom_jitter(aes(x = batch, y = value, col = batch))
```


```{r}
# Add the surrogate variables to the model matrix
mod_sva = lm(t(bottomly_edata) ~ as.factor(bottomly_metadata$strain) + sva.output$sv)
tidy_mod_sva <- tidy(mod_sva)
```

```{r}
est_compare <- tibble(
  LinearModel = tidy_mod %>% filter(term == "as.factor(bottomly_metadata$strain)DBA/2J") %>% select("estimate") %>% unlist,
  ComBat = tidy_mod_combat %>% filter(term == "as.factor(bottomly_metadata$strain)DBA/2J") %>% select("estimate") %>% unlist,
  SVA = tidy_mod_sva %>% filter(term == "as.factor(bottomly_metadata$strain)DBA/2J") %>% select("estimate") %>% unlist)
```

```{r}
png("Guzek_problem5a.png", width = 1500, height = 500)
lmcb <- ggplot(est_compare, aes(x=LinearModel, y=ComBat)) +
     geom_point(col = "darkgrey", alpha = .5, size = .5) + geom_abline(intercept = 0, slope = 1, col="darkorange") + geom_smooth(method = "lm")

lmsva <- ggplot(est_compare, aes(x=LinearModel, y=SVA)) +
     geom_point(col = "darkgrey", alpha = .5, size = .5) + geom_abline(intercept = 0, slope = 1, col="darkorange") + geom_smooth(method = "lm")

cbsva <- ggplot(est_compare, aes(x=ComBat, y=SVA)) +
     geom_point(col = "darkgrey", alpha = .5, size = .5) + geom_abline(intercept = 0, slope = 1, col="darkorange") + geom_smooth(method = "lm")
figure <- ggarrange(lmcb, lmsva, cbsva, labels = c("Linear Model vs. ComBat", "Linear Model vs. SVA", "ComBat vs. SVA"), nrow = 1, ncol = 3) + labs(title = "Guzek_problem5a")
annotate_figure(figure, top = text_grob("Guzek_problem5a"))
dev.off()
figure
```

```{r}
p_vals <- tibble(
  LinearModel = tidy_mod %>% filter(term == "as.factor(bottomly_metadata$strain)DBA/2J") %>% select("p.value") %>% unlist,
  ComBat = tidy_mod_combat %>% filter(term == "as.factor(bottomly_metadata$strain)DBA/2J") %>% select("p.value") %>% unlist,
  SVA = tidy_mod_sva %>% filter(term == "as.factor(bottomly_metadata$strain)DBA/2J") %>% select("p.value") %>% unlist)

#png("Guzek_problem5b.png", width = 1400, height = 500)
p_vals.gather <- gather(p_vals)
ggplot(p_vals.gather, aes(x = value)) + geom_histogram(fill = "darkorange") + facet_wrap(~key)
#dev.off()
```